Predicting Success of Motivational Interviews with Recurrent Neural Network and Probabilistic Model for Patient-Provider Communication Sequences

### 2017-08-31 paper-amia-summit-2018/amia_summit_2018.tex
update results

\textbf {Recurrent Neural Network (RNN)} is a type of Neural Network, which can send feedback from the current hidden state to the hidden state of the next time step. In this way, RNN can capture long-term dependencies for predicting the future events, which is the main advantage of RNN over feedforward Neural Network, MC, and HMM. The capability of remembering past event is very useful in motivational interviews where a behavior observed at some points in the interview is very informative for the future behaviors that will be observed. However, it was observed that RNN fails to capture long-term dependencies due to vanishing gradient problem \cite{bengio1993problem}. In order to mitigate this problem, Hochreiter et al.\cite{hochreiter1997long} proposed a special kind of RNN called Long Short Term Memory networks or simply LSTM. Later, the model was enhanced by including forget gates and the tanh activation function \cite{graves2013speech}. There are several variants of LSTM model, where a dramatic variation on the LSTM is the Gated Recurrent Unit or simply GRU \cite{cho2014properties}. GRUs are simpler than LSTM units and experimentally outperform than other models in many cases. We reiterate the mathematical formulation of GRU defined by Chung et al.\cite{chung2014empirical} as follows (Eq.~\ref{eq:firstgru}~to~\ref{eq:lastgru}):
\begin{equation}
z_t = \sigma(W_zx_t + U_zh_{t-1} + b_z)
\label{eq:firstgru}
\end{equation}
\begin{equation}
r_t = \sigma(W_rx_t + U_rh_{t-1} + b_r)
\end{equation}
\begin{equation}
\tilde h_t = tanh(W_hx_t + r_t \odot U_hh_{t-1} + b_h) 
\end{equation}
\begin{equation}
h_t = z_t \odot h_{t-1} + (1-z_t) \odot \tilde h_t
\label{eq:lastgru}
\end{equation}  
\begin{equation}
total\ loss = \alpha \cdot \frac{1}{T}\sum_{t=1}^T loss(\bar y^{(t)},y^{(t)}) + (1 - \alpha) \cdot loss(\bar y^{(T)},y^{(T)})
\label{eq:loss}
\end{equation}  
\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.90\textwidth]{figures/rnn.eps}
    \caption{\textbf{Proposed RNN model with target replication}.}
    \label{fig:rnn-model}
\end{figure}
The architecture of the proposed Recurrent Neural Network is shown in Figure~\ref{fig:rnn-model}. We used one hidden LSTM or GRU layer of 20 nodes after embedding the given myscope codes from the input sequence. Then, we applied a softmax layer at each time step for predicting the target. This target replication strategy was used in several studies \cite{lipton2015learning,choi2016doctor} after introduced by Lee et al.\cite{lee2015deeply} in 2015 for convolutional neural networks. As it can be seen from equation~(\ref{eq:loss}), our loss is the combination of final loss and the mean of the losses over all sequence steps, where T is the total number of sequence step, $\bar y$ is the output at step t, and $\alpha\ \epsilon\ [0, 1]$ is a hyperparameter that determines the relative importance of intermediate outputs. LSTM and GRU are trained on 70\% of the data and validated on 20\%. The remaining 10\% of the data is used as a test set for reporting the performance of the model. Our models contain several hyperparameters such as embedding dimension, number of hidden units, learning rate, batch size, etc., which were determined by the validation set. We implemented our models in Tensorflow with adam optimizer and early stopping based on the validation loss. Experimentally, we observed that our model converges to optimal after 100 epoch and have no effect of dropout and regularization. The code for all models is publicly available at https://github.com/teanalab/myscope-sequential-analysis.

\subsection*{\textit{Evaluation metrics}}
Performance of the proposed method was evaluated in terms of precision, recall, and F-measure using 10 folds cross-validation and weighted macro-averaging of these metrics over the folds. A true positive (TP) was counted when the method correctly classified a sequence into its actual class; a false positive (FP) was counted for a class when the method incorrectly classified a sequence into that class; a false negative (FN) for an actual class of sequence was counted when the method incorrectly classified the sequence into other class. The Precision of a class was defined as the ratio of the numbers of correctly classified sequences and all sequences identified as belonging to that particular class by the classifier (i.e Precision = TP / (TP + FP)). Recall of a class was defined as the ratio between the numbers of correctly classified sequences and all sequences of that particular class in the gold standard (i.e. Recall = TP / (TP + FN)). F-measure is computed as the harmonic mean of precision and recall (i.e. 2 x Precision x Recall / (Precision + Recall)). However, Accuracy is simply computed as the ratio of correctly classified sequences and a total number of sequences.

@@

\begin{table}[h]
\centering
\caption{Performance of MC and HMM for predicting success of natural patient-provider communication sequences. The largest value for each performance metric is highlighted in bold.}
\label{tab:result_under_sampled}
  \begin{tabular}{|l|l|l|l|l|l|}
  \hline
   \textbf{Method} & \textbf{Accuracy}  & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Score}\\ \hline

@@

\subsection*{\textit{Predictive performance for under-sampled PPC code sequences}}
Predictive performance summary of the proposed methods on natural sequences is presented in Table~\ref{tab:result_under_sampled}. Three major conclusions can be drawn from the results in Table~\ref{tab:result_under_sampled}. First, HMM-based method generally outperforms MC-based one across all metrics for natural sequences. Second, first-order HMM has the best performance in terms of precision (0.7980) and F-measure (0.7989), while second-order HMM achieves the highest recall (0.8449). Third,  it follows that second-order MC and HMM have lower precision and F-measure, but higher recall than first-order MC and HMM. In particular, precision and F-measure decrease by 6.74\% and 2.39\%, in case of MC, and by 7.27\% \& 2.09\%, in case of HMM, when going from first- to second-order models. In contrast, recall improves by 3\%--5\%, in case of both MC and HMM, when going from first- to second-order models. \\

\begin{table}[h]
\centering
\caption{Performance of MC and HMM for predicting success of natural patient-provider communication sequences. The largest value for each performance metric is highlighted in bold.}
\label{tab:result_over_sampled}
  \begin{tabular}{|l|l|l|l|l|l|}
  \hline
   \textbf{Method} & \textbf{Accuracy}  & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Score}\\ \hline

@@

\subsection*{\textit{Predictive performance for over-sampled PPC code sequences}}
Table~\ref{tab:result_over_sampled} summarizes predictive performance of the proposed methods on alternating sequences. These results indicate that second-order HMM had better recall than MC, achieving 97.13\% recall. Similar to the case of natural sequences, prediction method based on second-order HMM has the highest recall of 0.9713 among all models. However, different from natural sequences, second-order MC has the highest precision (0.9778) as well as the highest F-measure (0.9736). We also observed that second-order MC and HMM outperform first-order MC and HMM. Specifically, moving from first-order to second-order MC improves precision, recall and F-measure of the MI success prediction method by 0.01\%, 2.72\% and 1.37\%, respectively. Moving from first-order to second-order HMM improves precision, recall and F-measure of the MI success prediction method by 3.32\%, 4.30\% and 3.70\%, respectively.

>>>>

\textbf {Recurrent Neural Network (RNN)} is a type of Neural Network, which can send feedback from the current hidden state to the hidden state of the next time step. In this way, RNN can capture long-term dependencies for predicting the future events, which is the main advantage of RNN over feedforward Neural Network, MC, and HMM. The capability of remembering past event is very useful in motivational interviews where a behavior observed at some points in the interview is very informative for the future behaviors that will be observed. However, it was observed that RNN fails to capture long-term dependencies due to vanishing gradient problem \cite{bengio1993problem}. In order to mitigate this problem, Hochreiter et al.\cite{hochreiter1997long} proposed a special kind of RNN called Long Short Term Memory networks or simply LSTM. Later, the model was enhanced by including forget gates and the tanh activation function \cite{graves2013speech}. There are several variants of LSTM model, where a dramatic variation on the LSTM is the Gated Recurrent Unit or simply GRU \cite{cho2014properties}. GRUs are simpler than LSTM units and experimentally outperform than other models in many cases. We reiterate the mathematical formulation of GRU defined by Chung et al.\cite{chung2014empirical} as follows (Eq.~\ref{eq:firstgru}~to~\ref{eq:lastgru}):
\begin{equation}
z_t = \sigma(W_zx_t + U_zh_{t-1} + b_z)
\label{eq:firstgru}
\end{equation}
\begin{equation}
r_t = \sigma(W_rx_t + U_rh_{t-1} + b_r)
\end{equation}
\begin{equation}
\tilde h_t = tanh(W_hx_t + r_t \odot U_hh_{t-1} + b_h) 
\end{equation}
\begin{equation}
h_t = z_t \odot h_{t-1} + (1-z_t) \odot \tilde h_t
\label{eq:lastgru}
\end{equation}  
\begin{equation}
total\ loss = \alpha \cdot \frac{1}{T}\sum_{t=1}^T loss(\bar y^{(t)},y^{(t)}) + (1 - \alpha) \cdot loss(\bar y^{(T)},y^{(T)})
\label{eq:loss}
\end{equation}  
\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.90\textwidth]{figures/rnn.eps}
    \caption{\textbf{Proposed RNN model with target replication}.}
    \label{fig:rnn-model}
\end{figure}
The architecture of the proposed Recurrent Neural Network is shown in Figure~\ref{fig:rnn-model}. We used one hidden LSTM or GRU layer of 20 nodes after embedding the given myscope codes from the input sequence. Then, we applied a softmax layer at each time step for predicting the target. This target replication strategy was used in several studies \cite{lipton2015learning,choi2016doctor} after introduced by Lee et al.\cite{lee2015deeply} in 2015 for convolutional neural networks. As it can be seen from equation~(\ref{eq:loss}), our loss is the combination of final loss and the mean of the losses over all sequence steps, where T is the total number of sequence step, $\bar y$ is the output at step t, and $\alpha\ \epsilon\ [0, 1]$ is a hyperparameter that determines the relative importance of intermediate outputs. Our model contains several hyperparameters such as embedding dimension, number of hidden units, learning rate, batch size, etc., which were determined by the validation set. We implemented our models in Tensorflow with adam optimizer and early stopping based on the validation loss. Experimentally, we observed that our model converges to optimal after 100 epoch and have no effect of dropout and regularization. The code for all models is publicly available at https://github.com/teanalab/myscope-sequential-analysis.

\subsection*{\textit{Evaluation metrics}}
Performance of the proposed method was evaluated in terms of precision, recall, and F-measure using 10 folds cross-validation and weighted macro-averaging of these metrics over the folds. However, LSTM and GRU are trained on 70\% of the data and validated on 20\%. The remaining 10\% of the data is used as a test set for reporting the performance of the model. A true positive (TP) was counted when the method correctly classified a sequence into its actual class; a false positive (FP) was counted for a class when the method incorrectly classified a sequence into that class; a false negative (FN) for an actual class of sequence was counted when the method incorrectly classified the sequence into other class. The Precision of a class was defined as the ratio of the numbers of correctly classified sequences and all sequences identified as belonging to that particular class by the classifier (i.e Precision = TP / (TP + FP)). Recall of a class was defined as the ratio between the numbers of correctly classified sequences and all sequences of that particular class in the gold standard (i.e. Recall = TP / (TP + FN)). F-measure is computed as the harmonic mean of precision and recall (i.e. 2 x Precision x Recall / (Precision + Recall)). However, Accuracy is simply computed as the ratio of correctly classified sequences and a total number of sequences.

@@

\begin{table}[h]
\centering
\caption{Performance of MC, HMM, and RNN for predicting success of under-sampled patient-provider communication sequences. The highest value for each performance metric is highlighted in bold.}
\label{tab:result_under_sampled}
  \begin{tabular}{|l|l|l|l|l|l|}
  \hline
   \textbf{Method} & \textbf{Accuracy}  & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Score}\\ \hline

@@

\subsection*{\textit{Predictive performance for under-sampled PPC code sequences}}
We applied small learning rate 0.0005 and 150 epoch with early stopping strategy for training deep learning models when small size of under-sampled data is used. Predictive performance summary of the proposed methods on under-sampled sequences is presented in Table~\ref{tab:result_under_sampled}. Five major conclusions can be drawn from the results in Table~\ref{tab:result_under_sampled}. First, recurrent neural networks outperforms over probabilistic models, and achieved 18.50\%-39.88\% higher F1-score. Second, GRU with target replication has the best performance over all other RNN methods, and achieved accuracy 0.8858, F1-score 0.8853 with precision 0.8917 and recall 0.8858. Third, target replication strategy improves the performance of GRU and LSTM, while conventional GRU shows better performance than LSTM with target replication. Fourth, among probabilistic models, MC-based method generally outperforms HMM across all metrics for under-sampled sequences. Fifth, it follows that second-order MC have lower acuracy, precision, recall and F-measure than first-order MC. In particular, acuracy, precision, recall and F-measure decrease by 15.80\%, 15.92\%, 15.80\% and 15.89\%, when going from first- to second-order models. At last, accuracy and recall illustrate exactly the same values for all models. \\

\begin{table}[h]
\centering
\caption{Performance of MC, HMM, and RNN for predicting success of over-sampled patient-provider communication sequences. The highest value for each performance metric is highlighted in bold.}
\label{tab:result_over_sampled}
  \begin{tabular}{|l|l|l|l|l|l|}
  \hline
   \textbf{Method} & \textbf{Accuracy}  & \textbf{Precision}  & \textbf{Recall} & \textbf{F1-Score}\\ \hline

@@

\subsection*{\textit{Predictive performance for over-sampled PPC code sequences}}
For over-sampled data, RNN models are trained for 100 epoch with early stopping strategy. Table~\ref{tab:result_over_sampled} summarizes predictive performance of the proposed methods on over-sampled sequences. These results indicate that HMM had better performance than second-order MC, achieving 5.74\% higher F1-score, while HMM still have 2.41\% lower F1-score than first-order Markov chain. Similar to the case of under-sampled sequences, target replication improves the performance of RNN models, and GRU with target replication has the highest F1-score of 0.8716 among all models. However, different from under-sampled sequences, predictive performance of probabilistic models increases whereas recurrent neural networks performance decreases. We also observed that all models have largest value in precision compared to all other performance metrics.
